---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# SMPD4

<!-- badges: start -->
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
<!-- badges: end -->

The goal of SMPD4 is to provide analysis ready data sets on SMPD4 data.  

Sphingomyelin Phosphodiesterase 4 (SMPD4) is a type of sphinogmyelinase protein
encoded by the SMPD4 gene. Patients with a bi-allelic loss of function variants
in this gene are associated with a condition known as neurodevelopmental disorder 
with microcephaly, arthrogryposis, and structural brain anomalies 
(NEDMABA; [618622](https://omim.org/entry/618622)).  

Few cases exist in the scientific literature with [Magini et al. (2019)](https://doi.org/10.1016/j.ajhg.2019.08.006) 
reporting 23 members of 12 unrelated families. The genotype-phenotype relationship
of NEDMABA is heterogeneous and aside from case-reports, little data is available
for statistical analysis. Magini et al. (2019) offer a supplementary table with
summarised clinical phenotype data of the included subjects. Unfortunately these
data are non standardised, heterogeneous and largely textual based. 

Further isolated examples exist in other papers and case studies, which have 
also been included (Bijarnia-Mahay et al. (2022), Ji et al. (2022), Monies et al. (2019), Ravenscroft et al. (2020)).  

This project aims to standardise and process this data into tidy format [(Wickham (2014))](https://doi.org/10.1016/j.ajhg.2019.08.006)
so it is analysis-ready and extensible for other researchers to better understand
the relationship between clinical factors discovered in the literature. 



## Installation

You can install the development version of SMPD4 from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("deanmarchiori/SMPD4")
```

## Example  

```{r}
library(SMPD4)

data("magini_2019")
head(magini_2019[,1:20])
```


## References

Bijarnia-Mahay, S., Somashekar, P. H., Kaur, P., Kulshrestha, S., Ramprasad, V. L., Murugan, S., ... & Shukla, A. (2022). Growth and neurodevelopmental disorder with arthrogryposis, microcephaly and structural brain anomalies caused by Bi-allelic partial deletion of SMPD4 gene. Journal of human genetics, 67(3), 133-136.

Magini, P., Smits, D. J., Vandervore, L., Schot, R., Columbaro, M., Kasteleijn, E., ... & Mancini, G. M. (2019). Loss of SMPD4 causes a developmental disorder characterized by microcephaly and congenital arthrogryposis. The American Journal of Human Genetics, 105(4), 689-705.

Ji, W., Kong, X., Yin, H., Xu, J., & Wang, X. (2022). Case Report: Novel Biallelic Null Variants of SMPD4 Confirm Its Involvement in Neurodevelopmental Disorder With Microcephaly, Arthrogryposis, and Structural Brain Anomalies. Frontiers in Genetics, 13.

Ravenscroft, G., Clayton, J. S., Faiz, F., Sivadorai, P., Milnes, D., Cincotta, R., ... & Davis, M. R. (2021). Neurogenetic fetal akinesia and arthrogryposis: genetics, expanding genotype-phenotypes and functional genomics. Journal of medical genetics, 58(9), 609-618.

Monies, D., Abouelhoda, M., Assoum, M., Moghrabi, N., Rafiullah, R., Almontashiri, N., ... & Alkuraya, F. S. (2019). Lessons learned from large-scale, first-tier clinical exome sequencing in a highly consanguineous population. The American Journal of Human Genetics, 104(6), 1182-1201.  

Wickham, H. . (2014). Tidy Data. Journal of Statistical Software, 59(10), 1–23. https://doi.org/10.18637/jss.v059.i10



## Data use and licencing  

These data are Copyright 2019 American Society of Human Genetics. Data is used under Elsevier user license https://www.elsevier.com/about/policies/open-access-licenses/elsevier-user-license

Use permitted for read, download, text and data mining. Used for Non-Commercial purposes. All outputs of this R package are data mining related for non-commerical use. 