---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# SMPD4

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The goal of SMPD4 is to provide analysis ready data sets on SMPD4 data.  

Sphingomyelin Phosphodiesterase 4 (SMPD4) is a type of sphinogmyelinase protein
encoded by the SMPD4 gene. Patients with a bi-allelic loss of function variants
in this gene are associated with a condition known as neurodevelopmental disorder 
with microcephaly, arthrogryposis, and structural brain anomalies 
(NEDMABA; [618622](https://omim.org/entry/618622)).  

Few cases exist in the scientific literature with [Magini et al. (2019)](https://doi.org/10.1016/j.ajhg.2019.08.006) 
reporting 23 members of 12 unrelated families. The genotype-phenotype relationship
of NEDMABA is heterogeneous and aside from case-reports, little data is available
for statistical analysis. Magini et al. (2019) offer a supplementary table with
summarised clinical phenotype data of the included subjects. Unfortunately these
data are non standardised, heterogeneous and largely textual based. This project
aims to standardise and process this data into tidy format [(Wickham (2014))](https://doi.org/10.1016/j.ajhg.2019.08.006)
so it is analysis-ready and extensible for other researchers to better understand
the relationship between clinical factors discovered in the literature. 



## Installation

You can install the development version of SMPD4 from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("deanmarchiori/SMPD4")
```

## Example  

```{r}
library(SMPD4)

data("smpd4_phenotype")
head(smpd4_phenotype[,1:20])
```


## References

Magini, P., Smits, D. J., Vandervore, L., Schot, R., Columbaro, M., Kasteleijn, E., ... & Mancini, G. M. (2019). Loss of SMPD4 causes a developmental disorder characterized by microcephaly and congenital arthrogryposis. The American Journal of Human Genetics, 105(4), 689-705. https://doi.org/10.1016/j.ajhg.2019.08.006  

Wickham, H. . (2014). Tidy Data. Journal of Statistical Software, 59(10), 1–23. https://doi.org/10.18637/jss.v059.i10



# Data use and licencing  

These data are Copyright 2019 American Society of Human Genetics. 

Data is used under Elsevier user license https://www.elsevier.com/about/policies/open-access-licenses/elsevier-user-license

Use permitted for read, download, text and data mining. Used for Non-Commercial purposes. 

All outputs of this R package are data mining related for non-commerical use. 

DOI: https://doi.org/10.1016/j.ajhg.2019.08.006
